### 2010-10-22 Thesis/Chapter1.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Virtually everyone knows this situation where a close relative has to go to the hospital, either for a simple routine check or a serious surgery. Hospitals are meant to be reinsuring, a refuge where people's health improves, where people get fixed. Obviously, patient safety is a critical concern in the medical industry. Yet, patients have been known to suffer injuries or even death due to errors of judgement or a lack of care and training. As an example, a report published in 2000 by the Institute of Medicine in the United States \citep{Kohn00} described two studies carried out in the 1990s based on large samples of hospital admissions. They found that the proportion of hospital admissions experiencing an adverse event, defined as causes of injuries by medical management, were 2.9 and $3.7\,$\%, respectively. The proportion of these events attributable to errors (that is, preventable adverse events, in other words medical errors) was 58 and $53\,$\%, respectively. Even when using the lower estimate, deaths due to preventable adverse events in 1997 in the US ($44\,000$) exceeded the deaths attributable to motor vehicle accidents ($43\,458$), breast cancer ($42\,297$) or AIDS ($16\,516$). Because many errors go unreported, detailed statistics on medical errors are fairly scarce (hospitals usually do not attempt to emphasize errors). Nevertheless, the problem is real. And this is not surprising, elaborate skills have to be mastered in the medical field. Moreover, achieving a high degree of personal competence is sometimes not sufficient as various members of staff must learn how to work as a team. In fact, one of the recommendations of this report \citep{Kohn00} was to \emph{establish interdisciplinary team training programs for providers that incorporate proven methods of training, such as simulation}. However, dispensing an appropriate training in such a complex and hazardous environment is very challenging. Because healthcare is a high risk industry (like aviation or military), training in the real world is too costly and dangerous. Consequently, various approaches were applied for teaching and training of medical practicians over history of healthcare.

@@

The most basic form of medical simulators are simple models of human anatomy. Hundreds of years ago, representations in clay and stone were already used to demonstrate clinical features of disease states and their effects on humans. Models have been found from many cultures and continents. Nowadays, similar passive models are still used to help students learn the anatomy. In medical schools, live animals are also used for teaching physiology and surgery classes. Dogs, pigs, goats and sheep are commonly used in medical training. The exercices essentially consist in anesthetising the animal and practicing surgical techniques. They may also be severely injured to teach participants how to treat and manage various traumatic injuries. If animal training allows a better realism than passive models, this type of training is being progressively abandoned because of both the prohibitive cost and obvious ethical reasons. Alternatives include active models of human patients that attempt to reproduce living anatomy or physiology. The first example was created in the early 1960s by Asmund Laerdal for training in mouth to mouth ventilation. This simulator of a dying victim not breathing and lacking a heart beat, became known as called Resusci-Anne and had been widely used for CPR training thanks to an internal spring attached to the chest wall \citep{Cooper04}. In the mid-1960s, Sim One became the first mannequin controlled by a computer. The chest was moved with breathing, the eyes blinked and the pupils could dilate for instance. But the computer technology was too expensive for commercialisation at the time and only one mannequin was built. Over time, computerised mannequins have substantially improved and today such systems are essentially integrated into training centers that aim at recreating the operating room environment. Scenarios can be created to expose medical professionals to diverse clinical situations with the goal of improving the communication between staff members. Rare but critical clinical situations that may require specialised training can also be simulated. Another approach to medical training, that may also be combined with computerised mannequins, is the use of computer-based simulators. Such systems provide an elegant solution for medical training since they can offer a realistic and configurable training environment. Computer-based simulators appear as an ideal tool to assess students' skills since virtually every parameter of the simulation can be measured and recorded. Increasing computational power and growing interest for the field over the last decade led to the rapid development of reasonably sophisticated simulation systems. Yet, computer-based simulators still face numerous challenges.

@@

While substantial progress was accomplished during the past ten years, much remains to be done. In fact, the research described in this PhD fits in this context of computed-based simulation. Medical simulation can be broken up into different areas of research, all are very challenging.
%
\begin{description}
\item[Photorealistic rendering:] first, the scene needs to look real. A rendered image can be understood in terms of a number of visible features that we want to reproduce in a simulated image: colors, brightness, shadows, reflection, transparency, refraction, diffraction, etc. Therefore, a series of fairly complex calculations is required for all visible structures within each image (usually displayed at $30\,$Hz). 
\item[Collision detection] involves algorithms to check for collision between solids. Without them, tools could go through organs and other obstacles during the simulation. The collision detection algorithm has to find all intersections between the objects at each time step of the simulation. 
\item[Contact modelling:] once the collision between two objects has been detected, a variety of responses to this contact may be formulated. The two solids may rebound away from each other, slide (with or without friction), or settle into relative static contact. This step must also be carried out at each time step. 
\item[Anatomical structure modelling and haptic feedback:] the simulator must also take into account the physics involved in the deformation of anatomical structures. Indeed, when a pressure is exerted on a given organ, the latter must deform and react according to its mechanical properties. The force created by this interaction may be sent back to the user through a so-called haptic device. This device provides the user with a sense of touch so he can feel the forces and vibrations due to the contact. The force feedback must take place at a very high frequency (often between $ 500 $ and $ 1\,000\,$Hz) for a smooth sense of touch. 
\end{description}
%
Although each of these steps alone is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Real-time means that one second of simulation corresponds to one second of real life. Of particular interest to us in this PhD is the accurate modelling of the deformation of anatomical structures in real-time.

Along with the increase in computational power, approaches for modelling anatomical structures became more elaborate and less approximative. These different approaches may be classified into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods based on the equations of continuum mechanics. Techniques based on geometry tolerate a degradation of realism as long as the result looks realistic and are typically used in the game industry. As an example, the technique of shape matching proposed by \cite{Muller05} makes use of a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. But the most popular technique applied in simulation is the mass-spring system \citep{Terzopoulos91,Provot95,Mosegaard05b} . In this method, the geometry is described by a network of masses connected together by springs. The system of equations is solved using Newton's second law. Finally, the most computationally demanding techniques are the one relying on the equations of continuum mechanics. These equations are regarded as the most accurate mathematical description available for modelling the deformation of a continuum. The finite element method is a numerical procedure that allows to solve the partial differential equations provided by continuum mechanics \citep{Cotin99}. With the continual growth in computational power, finite element formulations applied to medical simulation become more elaborate \citep{Sagar94,Felippa00,Debunne01,Miller07}. For instance, they tend to employ more accurate constitutive laws to better describe the complex mechanical behaviour of organs \citep{Fung93}.

>>>>

Virtually everyone knows this situation where a close relative has to go to the hospital, either for a simple routine check or a serious surgery. Hospitals are meant to be re-insuring, a refuge where people's health improves, where people get fixed. Obviously, patient safety is a critical concern in the medical industry. Yet, patients have been known to suffer injuries or even death due to errors of judgement or a lack of care and training. As an example, a report published in 2000 by the Institute of Medicine in the United States \citep{Kohn00} described two studies carried out in the 1990s based on large samples of hospital admissions. They found that the proportion of hospital admissions experiencing an adverse event, defined as causes of injuries by medical management, were 2.9 and $3.7\,$\%, respectively. The proportion of these events attributable to errors (that is, preventable adverse events, in other words medical errors) was 58 and $53\,$\%, respectively. Even when using the lower estimate, deaths due to preventable adverse events in 1997 in the US ($44\,000$) exceeded the deaths attributable to motor vehicle accidents ($43\,458$), breast cancer ($42\,297$) or AIDS ($16\,516$). Because many errors go unreported, detailed statistics on medical errors are fairly scarce (hospitals usually do not attempt to emphasize errors). Nevertheless, the problem is real. And this is not surprising, elaborate skills have to be mastered in the medical field. Moreover, achieving a high degree of personal competence is sometimes not sufficient as various members of staff must learn how to work as a team. In fact, one of the recommendations of this report \citep{Kohn00} was to \emph{establish interdisciplinary team training programs for providers that incorporate proven methods of training, such as simulation}. However, dispensing an appropriate training in such a complex and hazardous environment is very challenging. Because healthcare is a high risk industry (like aviation or military), training in the real world is too costly and dangerous. Consequently, various approaches were applied for teaching and training of medical practicians over history of healthcare.

@@

The most basic form of medical simulators are simple models of human anatomy. Hundreds of years ago, representations in clay and stone were already used to demonstrate clinical features of disease states and their effects on humans. Models have been found from many cultures and continents. Nowadays, similar passive models are still used to help students learn the anatomy. In medical schools, live animals are also used for teaching physiology and surgery classes. Dogs, pigs, goats and sheep are commonly used in medical training. The exercises essentially consist in anaesthetising the animal and practicing surgical techniques. They may also be severely injured to teach participants how to treat and manage various traumatic injuries. If animal training allows a better realism than passive models, this type of training is being progressively abandoned because of both the prohibitive cost and obvious ethical reasons. Alternatives include active models of human patients that attempt to reproduce living anatomy or physiology. The first example was created in the early 1960s by Asmund Laerdal for training in mouth to mouth ventilation. This simulator of a dying victim not breathing and lacking a heart beat, became known as called Resusci-Anne and had been widely used for CPR training thanks to an internal spring attached to the chest wall \citep{Cooper04}. In the mid-1960s, Sim One became the first mannequin controlled by a computer. The chest was moved with breathing, the eyes blinked and the pupils could dilate for instance. But the computer technology was too expensive for commercialisation at the time and only one mannequin was built. Over time, computerised mannequins have substantially improved and today such systems are essentially integrated into training centres that aim at recreating the operating room environment. Scenarios can be created to expose medical professionals to diverse clinical situations with the goal of improving the communication between staff members. Rare but critical clinical situations that may require specialised training can also be simulated. Another approach to medical training, that may also be combined with computerised mannequins, is the use of computer-based simulators. Such systems provide an elegant solution for medical training since they can offer a realistic and configurable training environment. Computer-based simulators appear as an ideal tool to assess students' skills since virtually every parameter of the simulation can be measured and recorded. Increasing computational power and growing interest for the field over the last decade led to the rapid development of reasonably sophisticated simulation systems. Yet, computer-based simulators still face numerous challenges.

@@

While substantial progress was accomplished during the past ten years, much remains to be done. In fact, the research described in this PhD fits in this context of computed-based simulation. Medical simulation can be broken up into different areas of research, all are very challenging.
%
\begin{description}
\item[Photo-realistic rendering:] first, the scene needs to look real. A rendered image can be understood in terms of a number of visible features that we want to reproduce in a simulated image: colours, brightness, shadows, reflection, transparency, refraction, diffraction, etc. Therefore, a series of fairly complex calculations is required for all visible structures within each image (usually displayed at $30\,$Hz). 
\item[Collision detection] involves algorithms to check for collision between solids. Without them, tools could go through organs and other obstacles during the simulation. The collision detection algorithm has to find all intersections between the objects at each time step of the simulation. 
\item[Contact modelling:] once the collision between two objects has been detected, a variety of responses to this contact may be formulated. The two solids may rebound away from each other, slide (with or without friction), or settle into relative static contact. This step must also be carried out at each time step. 
\item[Anatomical structure modelling and haptic feedback:] the simulator must also take into account the physics involved in the deformation of anatomical structures. Indeed, when a pressure is exerted on a given organ, the latter must deform and react according to its mechanical properties. The force created by this interaction may be sent back to the user through a so-called haptic device. This device provides the user with a sense of touch so he can feel the forces and vibrations due to the contact. The force feedback must take place at a very high frequency (often between $ 500 $ and $ 1\,000\,$Hz) for a smooth sense of touch. 
\end{description}
%
Although each of these steps alone is computationally demanding, the whole simulation process must be computed in real-time or close to real-time. Real-time means that one second of simulation corresponds to one second of real life. In that regard, a lot of work remains to be done in all these areas of research. Along with the increase in computational power, medical simulators endeavour to achieve the highest degree of realism possible. Of particular interest to us in this PhD is the accurate modelling of the deformation of anatomical structures in real-time.

In the area of biomechanics, approaches for modelling anatomical structures became more elaborate and less approximative. These different approaches may be classified into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods based on the equations of continuum mechanics. Techniques based on geometry tolerate a degradation of realism as long as the result looks realistic and are typically used in the game industry. As an example, the technique of shape matching proposed by \cite{Muller05} makes use of a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. But the most popular technique applied in simulation is the mass-spring system \citep{Terzopoulos91,Provot95,Mosegaard05b}. In this method, the geometry is described by a network of masses connected together by springs. The system of equations is solved using Newton's second law. Finally, the most computationally demanding techniques are the one relying on the equations of continuum mechanics. These equations are regarded as the most accurate mathematical description available for modelling the deformation of a continuum. The finite element method is a numerical procedure that allows to solve the partial differential equations provided by continuum mechanics \citep{Cotin99}. With the continual growth in computational power, finite element formulations applied to medical simulation become more elaborate \citep{Sagar94,Felippa00,Debunne01,Miller07}. For instance, they tend to employ more accurate constitutive laws to better describe the complex mechanical behaviour of organs.

Yet, few biomechanical models can be computed in real-time and many approximations are still in use in medical simulation. Even though the mechanical behaviour of most organs is known to be non-linear, anisotropic and viscoelastic \citep{Fung93}, simple linear models are often employed in medical simulators. The resulting deformation of these organs is rather inaccurate and the force feedback provided to the user is not correct. Moreover, most of the proposed models are concerned with the mechanical behaviour of solid organs despite the fact that many thin and hollow anatomical structures may be found in the human body. To our knowledge, no model based on continuum mechanics was suggested in medical simulation to simulate the deformation of thin structures. Therefore, new algorithms are required for both type of structures. More elaborate models are suitable to better describe the complex behaviour of solid structures (with greater accuracy) and physics-based models are needed to model thin anatomical structures, all within the context of a real-time environment.

### 2010-10-22 Thesis/Chapter10.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Physics-based reconstruction using shell elements}
\label{chap10}
\begin{shortAbstract}
In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane and offplane forces (bending and twist). The validity of our approach has been demonstrated though comparison with theoretical results. It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These preliminary results are very encouraging and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements. This work was the object of one publication \citep{Comas2010b}.
\end{shortAbstract}

@@

Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved. Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry. Literature about volumic mesh generation algorithms is fairly dense. However, there are only a few that are concerned about the generation of meshes over curved surfaces. One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface. If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces. The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}. The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation. Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}. However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution, and therefore requires some knowledge about the problem in order to choose an effective tolerance. \cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature. While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.

@@

However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure. In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell. By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap9:eq-deflection} used in our co-rotational shell FEM.

@@

By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached. A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method. 
\begin{figure}
\centering
\includegraphics[width=13.5cm]{chapter10/exampleCylinder.pdf}
\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b). In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh. (d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
\label{chap10:fig-cylinder}
\end{figure}

@@

\subsubsection*{Meshing of anatomical structures.}
This approach has been applied to approximate more complex anatomical geometries with curved shell elements. Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap10:fig-liver}) and an aneurysm (\fig{chap10:fig-aneurysm}). In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box. 
%
\begin{figure}
\centering
\includegraphics[width=13.5cm]{chapter10/resultsLiver.pdf}
\caption {(a) the targeteted high resolution Glisson's capsule mesh ($8\,000$ triangles). (b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).} 
\label{chap10:fig-liver}
\end{figure}

@@

We proposed a framework for real-time modelling of thin anatomical structures. The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells. We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function. In summary, these polynomial shape functions are used in three different ways in our computational model:
%
\begin{description}
\item[(a)] to approximate complex geometrical shapes
\item[(b)] to compute internal forces via a co-rotational shell FEM
\item[(c)] to compute contact forces onto a curved triangle
\end{description}

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant). We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure.

\bigskip

The finite element procedure that we propose is largely developed based on physical understanding without the use of mathematical shell theories. Indeed, our shell FEM was obtained by simply superimposing a plane stress and bending energies. According to \cite{Chapelle03}, this type of approach has a limited level of accuracy. The effective study of shell structures clearly requires a deep physical understanding of shell structural behaviours. The complexity of the physical behaviours of shells requires advanced mathematical analyses from theories that are still in development. The interested reader may refer to the book of \cite{Chapelle03} for more details on shell theory. If the range of application of our shell formulation might be somewhat limited, we did not find out any inconsistencies through our tests. We also need to keep in mind that the accuracy required in medical simulation is not the same than the one demanded in structural analyses for instance. Moreover, the additional complexity of a shell finite element formulation based on true shell theory may also very well jeopardise our real-time constraints. Therefore, we do not think that the use of this simple shell formulation significantly diminishes our contribution to the field of medical simulation. Indeed, to our knowledge, our co-rotational shell FEM is the first description of a shell finite element model applied to simulating the deformation of thin anatomical structures.

Nevertheless, a few improvements are conceivable. Large strain could be allowed by using a non-linear strain tensor and more complex material laws (non-linear, viscoelasticity, etc.) could also be added to our shell formulation. However, those changes are substantial and their implementations are not straightforward. In addition, it is not clear how the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated). The answer may depend on the type of constraints applied on the object. Also (as already mentioned section \ref{chap9:GPU}), the real-time capacity of our algorithm could be further improved by implementing our shell FEM on GPU. Although an optimal parallel re-implementation is always challenging, we have already reviewed why there was no major obstacle to a GPU implementation.

Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach suffers from two major drawbacks. In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivised into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm. Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement may be very limited. The second drawbrack is that we also do not check the quality of the elements during our mesh process. If some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. Yet, meshing a surface with curved elements is a very complex problem and still an active area of research. In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

>>>>

\chapter{Physics-based reconstruction using shell elements}
\label{chap10}
\begin{shortAbstract}
In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach. This simple shell element can efficiently handle both membrane and off-plane forces (bending and twist). The validity of our approach has been demonstrated though comparison with theoretical results. It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation. These results are very encouraging and show the potential of such models. Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels. However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches. In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements.
\end{shortAbstract}

@@

Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved. Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry. Literature about volumetric mesh generation algorithms is fairly dense. However, there are only a few that are concerned about the generation of meshes over curved surfaces. One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface. If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces. The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}. The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation. Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}. However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution. Therefore it requires some knowledge about the problem in order to choose an effective tolerance. \cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature. While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.

@@

However, we need to realise that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure. In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell. By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap9:eq-deflection} used in our co-rotational shell FEM.

@@

By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached. A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method. 
\begin{figure}[ht]
\centering
\includegraphics[width=13.5cm]{chapter10/exampleCylinder.pdf}
\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b). In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh. (d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
\label{chap10:fig-cylinder}
\end{figure}

@@

\subsubsection*{Meshing of anatomical structures.}
This approach has been applied to approximate more complex anatomical geometries with curved shell elements. Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap10:fig-liver}) and an aneurysm (\fig{chap10:fig-aneurysm}). In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box. 
%
\begin{figure}[ht]
\centering
\includegraphics[width=13.5cm]{chapter10/resultsLiver.pdf}
\caption {(a) the targeteted high resolution Glisson's capsule mesh ($8\,000$ triangles). (b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).} 
\label{chap10:fig-liver}
\end{figure}

@@

Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach could be further improved. In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivided into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm. Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement is not very suitable. The second weakness is that we also do not check the quality of the elements during our mesh process. If some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. Yet, meshing a surface with curved elements is a very complex problem and still an active area of research. In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

\section{Conclusion}

Over the last two chapters, we proposed a complete framework for real-time modelling of thin anatomical structures. The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells. We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function. In summary, these polynomial shape functions are used in three different ways in our computational model:
%
\begin{description}
\item[(a)] to approximate complex geometrical shapes
\item[(b)] to compute internal forces via a co-rotational shell FEM
\item[(c)] to compute contact forces onto a curved triangle
\end{description}

We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy. The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant). We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure. This work on modelling thin anatomical structures was the object of two main publications \citep{Comas2010a,Comas2010b}.

### 2010-10-22 Thesis/Chapter11.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\subsubsection*{Solid structure modelling.}
We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA. This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA. 
%By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

@@

Regarding our geometric surface reconstruction based on shells, we were able to mesh complex anatomical structures with a fairly small number of shell elements. However, the main drawback of our method is the lack of any assessment of the quality of the elements created. Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. Yet, meshing a surface with curved elements is a very challenging problem and still an active area of research. In contrast to all geometrical techniques described in the literature, we presented a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

@@

One of the strongest constraint in medical simulation is the demand for real-time computations. This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photorealistic rendering etc.). In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures. The lack of accuracy induced by these simplifications is of great concern to us. In this PhD, a part ot the solution was to accelerate the computations by using the GPU. While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea. Firstly, a standard is yet to be imposed in GPGPU. While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless. In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance (up to twice as slow). Secondly, GPU programming is not simple to master. Although we have seen many improvements in flexibility of programmation over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code. As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least. But most importantly, GPU acceleration is only interesting if the code is highly parallelisable. The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties. Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps. Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).

>>>>

\subsubsection*{Solid structure modelling.}
We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA. This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our efficient re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA. 
%By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

@@

Regarding our geometric surface reconstruction based on shells, we were able to mesh complex anatomical structures with a fairly small number of shell elements. However, the main drawback of our method is the lack of any assessment of the quality of the elements created. Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing. Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements. Yet, meshing a surface with curved elements is a very challenging problem and still an active area of research. In contrast to all geometrical techniques described in the literature, we presented a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

@@

One of the strongest constraint in medical simulation is the demand for real-time computations. This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photo-realistic rendering etc.). In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures. The lack of accuracy induced by these simplifications is of great concern to us. In this PhD, a part ot the solution was to accelerate the computations by using the GPU. While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea. Firstly, a standard is yet to be imposed in GPGPU. While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless. In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance (up to twice as slow). Secondly, GPU programming is not simple to master. Although we have seen many improvements in flexibility of programming over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code. As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least. But most importantly, GPU acceleration is only interesting if the code is highly parallelisable. The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties. Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps. Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).
